Reviving the Lost Art of Combinator Graph Reduction

### 2023-01-24 experiments/superg/doc/paper/paper.tex
Stub out sections of preliminaries.

\section{Preliminaries}
\label{sec:prelim}
\textbf{TODO: write section. Currently contains only a few small fragments}
\subsection{Combinatory Logic}
Starting from a lambda calculus expression, we first translate it into an expression consisting exclusively of the application of a small set of predefined functions, or \textit{combinators}.
It turns out that any lambda calculus expression can be expressed using just three combinators:
\begin{itemize}
    \item $S \ f \ g \ x = f \ x \ (g \ x)$
    \item $K \ x \ y = x$
    \item $I \ x = x$\footnote{It turns out that $SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$, and it simplifies the example.}
\end{itemize}

@@

\subsection{Graph Reduction}
% In the reduction graph, vertices are combinators, constants or function applications, and edges connect the function and argument for application (combinators and constants have no outgoing edges, they are leaf nodes).

@@

We observe that our Miranda-style engine is more than twice as fast as the original Miranda.
While we have not attempted an in-depth evaluation of the performance differences between these two systems, we can point to two likely reasons that our implementation performs better on this benchmark:
\begin{itemize}
    \item Miranda supports (exclusively) arbitrary precision integers.
          They are implemented by the high-performance GNU MP~\cite{gnump} library, but are still slower than the native 32-bit machine integers that our engine uses.
    \item As mentioned in~\ref{sec:miranda}, the original Miranda must be compiled with optimizations disabled.
          Without these optimizations, the compiler is likely to generate a slower implementation of key functions.
\end{itemize}

@@

\subsection{SIMD Combinator Reduction}
In his talk \emph{Combinator Revisited}~\cite{kmett_combinators_2018}, Kmett hypothesizes the use of SIMD instructions to perform combinator reductions.
The rationale is that combinators have a relatively simple implementation, and SIMD instructions are getting more and more powerful, so potentially they can be used to implement the reduction of multiple combinators in a single instructions.
Assuming the reduction of a single combinator uses 64-bit wide values, the same routine implemented with AVX-512 instructions could be used to perform as many as 8 reductions in parallel on a single core.
An eightfold improvement in performance is certainly promising, although it is unclear what level of parallelism a typical program may allow for.
A work-in-progress version of the compiler is available at \url{https://github.com/ekmett/coda}.

>>>>

\section{Preliminaries}
\label{sec:prelim}
\textbf{TODO: write section. Currently contains only a few small fragments}
\subsection{Lambda Calculus}

\subsection{Combinatory Logic}
Starting from a lambda calculus expression, we first translate it into an expression consisting exclusively of the application of a small set of predefined functions, or \textit{combinators}.
It turns out that any lambda calculus expression can be expressed using just three combinators:
\begin{itemize}
    \item $S \ f \ g \ x = f \ x \ (g \ x)$
    \item $K \ x \ y = x$
    \item $I \ x = x$\footnote{It turns out that $SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$, and it simplifies the example.}
\end{itemize}

@@

\subsection{Bracket abstraction}

@@

We observe that our Miranda-style engine is more than twice as fast as the original Miranda.
While we have not attempted an in-depth evaluation of the performance differences between these two systems, we can point to two likely reasons that our implementation performs better on this benchmark:
\begin{itemize}
    \item Miranda supports (exclusively) arbitrary precision integers.
          They are implemented by the high-performance GNU MP~\cite{gnump} library but are still slower than the native 32-bit machine integers that our engine uses.
    \item As mentioned in~\ref{sec:miranda}, the original Miranda must be compiled with optimizations disabled.
          Without these optimizations, the compiler is likely to generate a slower implementation of key functions.
\end{itemize}

@@

\subsection{SIMD Combinator Reduction}
In his talk \emph{Combinator Revisited}~\cite{kmett_combinators_2018}, Kmett hypothesizes the use of SIMD instructions to perform combinator reductions.
The rationale is that combinators have a relatively simple implementation, and SIMD instructions are getting more and more powerful, so potentially they can be used to implement the reduction of multiple combinators in a single instruction.
Assuming the reduction of a single combinator uses 64-bit wide values, the same routine implemented with AVX-512 instructions could be used to perform as many as 8 reductions in parallel on a single core.
An eightfold improvement in performance is certainly promising, although it is unclear what level of parallelism a typical program may allow for.
A work-in-progress version of the compiler is available at \url{https://github.com/ekmett/coda}.
